---
title: "FigSI"
author: "Zhenna Jiao"
date: "`r Sys.Date()`"
output: html_document
---
# Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(readxl)
library(ggpubr)
library(stringr)
library(readr)
library(Rmisc)
library(dplyr)
library(smoother)
library(reshape2)
library(matrixStats)
library(RColorBrewer)
library(ComplexHeatmap)
library(tidyr)
```

## Import the data directory

```{r}
fig_dir <- "~/methylation/project/EM_seq/analysis/technical/figure/fig2/"
proj <- "~/methylation/project/EM_seq/data/metadata/"
technical_dat_dir <- "~/methylation/project/EM_seq/analysis/technical/data/"
meta_df <- readRDS("~/methylation/project/EM_seq/data/metadata/meta_df.rds")
dat_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/result/technical/"
```

## FigS6 A-B  Make a violin plot for the  input concentration of EV-DNA (EV-DNA in patients' plasma) and the coverage of  the DNA sequencing of the genome 

```{r}
## import data
input_df <- read_excel(paste0(technical_dat_dir,"downsample.xlsx"),sheet = "downsample")

dat <- input_df[,c(1,19:20)]
rownames(dat) <- dat$sample_name...1
colnames(dat) <- c("smp","plasma_conc","coverage")
state <- dat$smp
dat$status <- substr(state,1,2)%>%as.factor()
dat$type <- c(rep("GP",21),rep("GC",23))%>%as.factor()
p_value_cov <- wilcox.test(dat$coverage[1:21],dat$coverage[22:44])[["p.value"]]

## Make a  plot for the difference between GP and GC  on the coverage of the genome  
p7 <- ggplot(dat, aes(x =type, y =coverage,col=type)) +
          geom_violin(aes(fill=type,alpha=0.5),scale="width")+
          scale_fill_manual(values=c("#FB8D8D","#C0E1FF"))+
          geom_boxplot(outlier.shape = NA,lwd=0.8,width=0.05)+
          geom_jitter(position=position_jitter(0.1),aes(fill=type))+
          scale_color_manual(values=c("#FB8D8D","#C0E1FF"))+
          theme_classic()+
          ylim(0,1)+
          labs(x ="", y = "Coverage Rate", 
                title = paste0("coverage rate between gp and gc ,the p.value is ",round(p_value_cov,3)))+
          theme(axis.text.x = element_text(size = 15,colour = "black"),
                axis.text.y = element_text(size = 15,colour = "black"),
                axis.title = element_text(size = 15,colour = "black"),
                axis.line = element_line(size=1,linetype = 1,color = "black"))
   
# ggsave(plot = p7,filename =paste0(fig_dir,"figS2F_1_downsample_sample_cov_20240902.pdf"),
#        width = 8,height = 6 ,dpi = 300)
        print(p7)

## Make a  plot for the difference between GP and GC on the input concentration of EV-DNA   
p_value_input <- wilcox.test(dat$plasma_conc[1:21],dat$plasma_conc[22:44])[["p.value"]]
p8 <- ggplot(dat, aes(x =type, y =plasma_conc,col=type)) +
          geom_violin(aes(fill=type,alpha=0.5),scale="width")+
          scale_fill_manual(values=c("#FB8D8D","#C0E1FF"))+
          geom_boxplot(outlier.shape = NA,lwd=0.8,width=0.05)+
          geom_jitter(position=position_jitter(0.1), aes(fill=type))+
          scale_color_manual(values=c("#FB8D8D","#C0E1FF"))+
          theme_classic()+
          labs(x =" ", y = "Input Concentration (ng/ml)", 
             title = paste0("input concentration between gp and gc ,
                            the p.value is ",round(p_value_input,3)))+
          theme(axis.text.x = element_text(size = 15,colour = "black"),
                axis.text.y = element_text(size = 15,colour = "black"),
                axis.title = element_text(size = 15,colour = "black"),
                axis.line = element_line(size=1,linetype = 1,color = "black"))
       
# ggsave(plot = p8,filename =paste0(fig_dir,"figS2F_downsample_sample_plasma_concentration_20240902.pdf"),
# width = 8,height = 6 ,dpi = 300)
print(p8)

```

## FigS7 A Bar plot showing the distribution of sample ages across control, early-stage, and advanced-stage groups.

```{r,fig.width=8,fig.height=6}
## import data
age_gender_df <- read_excel(paste0(technical_dat_dir,"age_and_gender.xlsx"),sheet = "filter_sample")
age_gender_df$Stage <- factor(age_gender_df$Stage, levels = c("Control", "Early Stage", "Advanced Stage"))
## do boxplot
p6 <-ggplot(age_gender_df, aes(x=Stage,y=Age))+
    geom_boxplot(outlier.shape = NA,lwd=0.6,
                 aes(fill=Stage),alpha=0.9)+
    geom_jitter(position=position_jitter(0.1), 
                aes(colour=Gender))+
    theme_classic()+
  labs(x = "", y = "Age",lwd = 1.5)+
  ylim(0,100)+
  scale_color_manual(values=c("#DE8686","#AACBFE"))+
  scale_fill_manual(values = c("#63A6D0","#F9A68A","#D34447"))+
  theme(axis.text.x = element_text(size = 15,colour = "black"),
        axis.text.y = element_text(size = 15,colour = "black"),
        axis.title = element_text(size = 15,colour = "black"),
        axis.line = element_line(size=1,linetype = 1,color = "black"),
          legend.text = element_text(size = 15), 
          legend.title = element_text(size = 15), 
          plot.title = element_text(size = 15),
          legend.position = "right"
        )
# ggsave(plot = p6,filename = paste0(fig_dir,"figS2E_boxplot_age_gender.pdf"),width = 8,height = 6,dpi = 300)
print(p6)
```

## FigS7 B  Methylation level variance  in 1-MB bins of the genome between GP and GC

```{r}
merge_df <- readRDS("~/methylation/project/EM_seq/analysis/technical/data/ds_1kb_count.rds")
mat2<- merge_df
meth<- merge_df
meth$idx <- floor(meth$start / 1000000)  # generate index for each 1,000,000 bp bin
meth[is.na(meth)]<-0
meth<-sapply(meth[,-1], as.numeric)
meth<-data.frame(meth)
meth$Chr<-mat2$chr
counts <- aggregate(meth[, grep("mC|nC|uC", colnames(meth))], by = list(meth$Chr, meth$idx), FUN=sum)  
# calculate methylation
for (indexmC in grep("_mC", colnames(counts))) { #index columns containing "_mC" in colname
  indexuC <- grep(gsub("_mC", "_uC", colnames(counts)[indexmC]), colnames(counts)) #index columns with uC
  m<-counts[,indexmC]/(counts[,indexuC]+counts[,indexmC])
  n<-ncol(counts)
  counts<-data.frame(counts,m)
  colnames(counts)[n+1]<-gsub("_mC", "_mod", colnames(counts)[indexmC]) 
}  
autosomes<-c("chr1$","chr2$","chr3$","chr4$","chr5$","chr6$","chr7$","chr8$","chr9$","chr10$",
             "chr11$","chr12$","chr13$","chr14$","chr15$","chr16$","chr17$","chr18$","chr19$","chr20$","chr21$","chr22$")

# select autosomes
counts<-counts[grep(paste(autosomes, collapse = "|"),counts$Group.1),]

mod_df=counts[,grep("_mod",colnames(counts))]
mod_df$pos<-paste(counts$Group.1, counts$Group.2, sep="_")
mod_df<-mod_df[complete.cases(mod_df),]
colnames(mod_df)<-gsub("_mod","",colnames(mod_df))

GP <-mod_df[,grep("gp", colnames(mod_df))]
rownames(GP)<-mod_df$pos

GC <-mod_df[,grep("sc", colnames(mod_df))]
rownames(GC)<-mod_df$pos
GP.var <-rowVars(as.matrix(GP))
GC.var<-rowVars(as.matrix(GC))
var.df<-data.frame(mod_df$pos,GP.var,GC.var)
var.df<-var.df%>%tidyr::separate(mod_df.pos,c("chr","pos"),"_")
var.melt<-melt(var.df,id.vars = c("chr","pos"))

chrOrder <-c("chr1","chr2","chr3","chr4","chr5","chr6","chr7","chr8","chr9","chr10","chr11",
             "chr12","chr13","chr14","chr15","chr16","chr17","chr18","chr19","chr20",
             "chr21","chr22")
var.melt$chr <- factor(var.melt$chr, chrOrder, ordered=TRUE) 
var.melt$status <- substr(var.melt$variable,1,2)
var.melt$status <- factor(var.melt$status,level=c("GP","GC"))

p5 <- ggplot(var.melt, aes(x =status, y =value)) +
        geom_boxplot(aes(fill=status),outlier.shape = NA,
                     lwd=0.7,alpha=0.8,width=0.5)+
        theme_classic()+
        ylim(0,0.001)+
        scale_fill_manual(values=c("#FB8D8D","#C0E1FF"))+
        labs(x = "", y = "Methylation variance",lwd = 1.5)+
        theme(
          axis.text.x = element_text(size = 15,hjust = 0.5,color = "black"),
        axis.text.y = element_text(size = 15,hjust = 0.5,color = "black"),
        axis.title.x = element_text(color = "black", size = 15),
        axis.title.y = element_text(color = "black", size = 15),
        axis.line = element_line(size=1,linetype = 1,color = "black"),
          legend.text = element_text(size = 15), 
          legend.title = element_text(size = 15), 
          plot.title = element_text(size = 15),
          legend.position = "none"
          )
print(p5)
```

## FigS7 C The distribution of %CpG modified level in 1-MB bins in different chromosomes

```{r}
## import data
merge_df <- readRDS("~/methylation/project/EM_seq/analysis/technical/data/ds_1kb_count.rds")
## Define the function to make a plot for the methylation level in 1-MB bins on different chromosomes 
chr_variance <- function(data,chromosome){
mat2 <- data
 
mat<-mat2%>%dplyr::filter(chr==chromosome)
# prepare 1-MB bins
ws<-1000
mat$bin<-c(1:nrow(mat))
mat<-data.frame(mat)
# index windows
for (i in grep ("bin", colnames(mat))){
  mat$index<-ceiling(mat[,i]/ws)
}

mat[is.na(mat)]<-0
mat<-sapply(mat[,-1], as.numeric)
mat<-data.frame(mat)
mat.100kb<-aggregate(mat[, grep("uC|mC|nC", colnames(mat))],
                     by = list(mat$index),
                     FUN = sum)

for (indexmC in grep("_mC", colnames(mat.100kb))) { #index columns containing "_mC" in colname
  indexuC <- grep(gsub("_mC", "_uC", colnames(mat.100kb)[indexmC]), colnames(mat.100kb)) #index columns with uC
  m<-mat.100kb[,indexmC]/(mat.100kb[,indexuC]+mat.100kb[,indexmC])
  n<-ncol(mat.100kb)
  mat.100kb<-data.frame(mat.100kb,m)
  colnames(mat.100kb)[n+1]<-gsub("_mC", "_mod", colnames(mat.100kb)[indexmC]) 
}  

mod_df=mat.100kb[,grep("_mod",colnames(mat.100kb))]
mod_df

#smooth
smooth<-data.frame(mat.100kb$Group.1)

for (i in 1:ncol(mod_df)){
  name<-colnames(mod_df)[i]
  tmp<-mod_df[,i]
  tmp.smooth<-smth.gaussian(tmp,
                            window = 10,
                            tails=TRUE,
                            alpha=1)
  n<-ncol(smooth)
  smooth<-data.frame(smooth,tmp.smooth)
  colnames(smooth)[n+1]<-name
}

healthy<-smooth[,grep("gp", colnames(smooth))]
healthy$bin<-c(1:nrow(healthy))
GC <-smooth[,grep("sc", colnames(smooth))]
GC$bin<-c(1:nrow(GC))

healthy.melt<-melt(healthy, id="bin")
hcc.melt<-melt(GC,id="bin")
colours=colorspace::sequential_hcl(30)
p1<-ggplot(healthy.melt, aes(x=bin, y=value, colour=variable))+
  geom_line()+
  theme_classic()+
  ggtitle("GP") +
  xlab(paste0("Postion on ",chromosome,"(x10^6)")) +
  ylab("Methylation value (%)")+
  ylim(0.5,0.9)+
  xlim(0,200)+
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, vjust = 0.5),
        axis.text.x = element_text(size = 15,  hjust = 0.5,color = "black"),
        axis.text.y = element_text(size = 15,hjust = 0.5,color = "black"),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(color = "black", size = 15),
        axis.title.y = element_text(color = "black", size = 15),
        axis.line = element_line(size=1,linetype = 1,color = "black"))
# ggsave(plot = p1,filename = paste0(fig_dir,"figS2C1_GP_chr4_meth_dist.pdf"),
#        width = 6,height = 4,dpi=300)
print(p1)

p2<-ggplot(hcc.melt, aes(x=bin, y=value, colour=variable))+
  geom_line()+
  theme_classic()+
  ggtitle("GC") +
  xlab(paste0("Postion on ",chromosome," (x10^6)")) +
  ylab("Methylation value (%)")+
  ylim(0.5,0.9)+
  xlim(0,200)+
  theme(legend.position = "none",
         plot.title = element_text(hjust = 0.5, vjust = 0.5),
        axis.text.x = element_text(size = 15,hjust = 0.5,color = "black"),
        axis.text.y = element_text(size = 15,hjust = 0.5,color = "black"),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(color = "black", size = 15),
        axis.title.y = element_text(color = "black", size = 15),
        axis.line = element_line(size=1,linetype = 1,color = "black"))
# ggsave(plot = p2,filename = paste0(fig_dir,"figS2C2_GC_chr4_meth_dist.pdf"),
#        width = 6,height = 4,dpi = 300)
print(p2)
}


## Plot the methylation level on chromosome 4
i=4
chr_variance(data = merge_df ,chromosome = paste0("chr",i))
```

## FigS7 D The distribution of %CpG modified level 

```{r}
## import data
sample_names <- meta_df$sample_list
smp_chr_data <- readRDS(paste0(dat_dir,"smp_chr_data.rds"))
mod_count <- readRDS(paste0(dat_dir,"mod_count.rds"))
df <- do.call(rbind, lapply(names(mod_count), function(name) mod_count[[name]]))
df$mod <- df$mod/100

df$bin<-NA
df$bin<-ifelse(df$mod>=0 & df$mod<=0.1,1,
               ifelse(df$mod>0.1 & df$mod<=0.2,2,
                      ifelse(df$mod>0.2 & df$mod<=0.3,3,
                             ifelse(df$mod>0.3 & df$mod<=0.4,4,
                                    ifelse(df$mod>0.4 & df$mod<=0.5,5,
                                           ifelse(df$mod>0.5 & df$mod<=0.6,6,
                                                  ifelse(df$mod>0.6 & df$mod<=0.7,7,
                                                         ifelse(df$mod>0.7 & df$mod<=0.8,8,
                                                                ifelse(df$mod>0.8 & df$mod<=0.9,9,
                                                                       ifelse(df$mod>0.9 & df$mod<=1,10,NA))))))))))

df$status<-NA
df$status[grep("gp",df$name)]<-"GP"
df$status[grep("sc",df$name)]<-"GC"
# df<-df[-grep("GP|GC", df$name),]
df$name<-as.character(df$name)
df<-aggregate(count ~ bin + name, FUN=sum, data=df)
sums<-aggregate(df$count, by=list(df$name), FUN=sum)
colnames(sums)<-c("name","total")
df <- merge(x = df, y = sums, by = "name")
df$freq<-(df$count/df$total)*100

df$status<-NA
df$status[grep("gp",df$name)]<-"GP"
df$status[grep("sc",df$name)]<-"GC"
# df$status[grep("Healthy_",df4$name)]<-"Healthy"

df2<-aggregate(freq ~ bin + status, FUN=mean, data=df)
df2$status<-factor(df2$status, levels=c("GP","GC"), ordered=TRUE) 

df$mean_freq<-NA
df3 <- df %>% right_join(df2, by=c("bin","status"))
df3$status<-factor(df3$status, levels=c("GP","GC"), ordered=TRUE)

## do histogram plot 
b.mod_gp <- ggplot(df3[df3$status == "GP",]) +
  geom_bar(aes(x=bin, y=freq.y), position="identity", stat="identity", color="black", fill="#C0E1FF") +
  geom_line(aes(x=bin, y=freq.x, colour=name), position="identity", stat="identity") +
  theme_classic() +
  ggtitle("GP")+
  ylab("% of modified CpG sites") +
  xlab("Modification level (%)")+
  # facet_wrap(~ status,scales = "free_x") +
  theme(legend.position = "none", 
        strip.text = element_text(face = "bold", size = 15, color = "black", angle = 0),
        strip.background = element_blank(), 
        axis.text.x = element_text(size = 15, angle = 90, hjust = 0.5,color = "black"),
        axis.text.y = element_text(size = 15,hjust = 0.5,color = "black"),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(color = "black", size = 15),
        axis.title.y = element_text(color = "black", size = 15),
        axis.line = element_line(size=1,linetype = 1,color = "black"),
        plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(1, 10, by = 1),
                      labels = c( "0-10", "10-20", "20-30", "30-40", 
                                  "40-50", "50-60", "60-70", "70-80", "80-90", "90-100"))

# ggsave(paste0(fig_dir,"figS2A_plot_GP_mod_level.pdf"),plot = b.mod_gp,width = 4,height =4,dpi = 300)
print(b.mod_gp)

b.mod_gc <- ggplot(df3[df3$status == "GC",]) +
  geom_bar(aes(x=bin, y=freq.y), 
           position="identity", stat="identity", color="black", fill="#FB8D8D") +
  geom_line(aes(x=bin, y=freq.x, colour=name), 
            position="identity", stat="identity") +
  theme_classic() +
  ylab("% of modified CpG sites") +
  xlab("Modification level (%)")+
  ggtitle("GC")+
  # facet_wrap(~ status,scales = "free_x") +
  theme(legend.position = "none", strip.text = element_text(face = "bold", size = 15, color = "black", angle = 0),
        strip.background = element_blank(), 
        axis.text.x = element_text(size = 15, angle = 90, hjust = 0.5,color = "black"),
        axis.text.y = element_text(size = 15,hjust = 0.5,color = "black"),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(color = "black", size = 15),
        axis.title.y = element_text(color = "black", size = 15),
        axis.line = element_line(size=1,linetype = 1,color = "black"),
        plot.title = element_text(hjust = 0.5)) +
        scale_x_continuous(breaks = seq(1, 10, by = 1),
                        labels = c( "0-10", "10-20", "20-30", "30-40",
                                    "40-50", "50-60", "60-70", "70-80", "80-90", "90-100"))


# ggsave(paste0(fig_dir,"figS2A_plot_GC_mod_level.pdf"),plot = b.mod_gc,width = 4,height =4,dpi = 300)
print(b.mod_gc)
```

## FigS8 A  Create a heatmap to visualize the correlation of methylation levels across DMRs detected between early and advanced stages.

```{r,fig.width=14,fig.height=10}
## set fig_stored dir
fig_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/figures/figS5-6/"
meta_scstage_df <- readRDS("~/methylation/project/EM_seq/data/metadata/meta_scstage_df.rds")
sc_dat_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/data/filter_sample/scstage_dmr_change/"
## import sc_satge data
scstage_DMR_meth <- readRDS(paste0(sc_dat_dir,"scstage_DMR_change_meth.rds"))
## calculate Pearson correlation coefficient matrix

data_keep_na <-  scstage_DMR_meth
row.names(data_keep_na) <- as.character(scstage_DMR_meth$chr_region)
data_keep_na <- data_keep_na[,-1]
colnames(data_keep_na) <- meta_scstage_df$smp
DMR_cor_keep_na <- cor(data_keep_na,use = "pairwise.complete.obs")
annotation_col_keepna <- data.frame(Stage=meta_scstage_df$stage,Gender=meta_scstage_df$gender)
DMR_cor_df_keepna <- as.data.frame(DMR_cor_keep_na)
rownames(annotation_col_keepna)  <- colnames(DMR_cor_df_keepna)

my_color=c("#4675FE","white","#F9A68A","#D34447")

gender_color=list(Gender=c(Male="#AACBFE",Female="#DE8686"))
stage_color=list(Stage=c("Early Stage"="#F9A68A","Advanced Stage"="#D34447"))

# pdf(paste0(fig_dir,"figS5A_heatmap_scstage_dmr_cor.pdf"),width = 14,height = 10)
pheatmap::pheatmap(DMR_cor_df_keepna,cluster_rows = T ,cluster_cols =T,show_rownames = T,show_colnames = T,
                   annotation_col = annotation_col_keepna,
                   annotation_colors = c(gender_color,stage_color),
                   scale = "none",
                  fontsize = 15,                
                  fontsize_row = 15,             
                  fontsize_col = 15,
                   cellwidth = 13,
                  cellheight =12,
                  color = colorRampPalette(my_color)(100))

# dev.off()
```

## FigS8 B Perform PCA analysis on DMRs called between the Early and Advanced Stages

```{r}
## import  data,  remove missing value
dat_rm_na <- scstage_DMR_meth%>%na.omit()
## Convert data format
rownames(dat_rm_na) <- as.character(dat_rm_na$chr_region)
dat_rm_na <- dat_rm_na[, -1]
tr_data <- t(dat_rm_na)
rownames(tr_data) <- colnames(dat_rm_na)
colnames(tr_data) <- rownames(dat_rm_na)
tr_data <- as.data.frame(tr_data)
tr_data$sample_stage <- meta_scstage_df$stage %>% as.factor()

## calculate PCs
tr.pca_all <- prcomp(tr_data[,1:(length(tr_data)-1)])

## do plot 
df1 <- tr.pca_all$x
pc_one <- as.data.frame(df1,tr_data$sample_stage)
pc_one$Stage <- tr_data$sample_stage
pc_one$smp_name <- substr(rownames(tr_data),1,4)
#Extract the variance contribution rate of the principal components and generate axis titles；
summ<-summary(tr.pca_all)
xlab<-paste0("PC1(",round(summ$importance[2,1]*100,2),"%)")
ylab<-paste0("PC2(",round(summ$importance[2,2]*100,2),"%)")
p1 <-ggplot(data = pc_one,aes(x=PC1,y=PC2,color=Stage))+
  geom_point(size=4.5)+
  labs(x=xlab,y=ylab)+
  theme_classic()+
  scale_color_manual(values = c("#D34447", "#F9A68A"))+
  theme(axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.text = element_text(size = 15,colour = "black"),
        axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
        axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
        axis.title = element_text(size = 15,colour = "black"))
 

# ggsave(paste0(fig_dir,"figS6_PCA_scstage_dmr_removena.pdf"),plot = p1,width = 8,height = 5,dpi = 300)
print(p1)
```

## FigS8 C 

```{r}
proj_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/result/annotation_dmr/"
dat_sc <- read_excel(paste0(proj_dir,"dmr_features.xlsx"),sheet = "scstage_dmr_feature_fisher")
dat_sc$logp <- -log(dat_sc$two_tail,10)
p4 <- ggplot(dat_sc, aes(y= elments, x= logp)) +
      geom_bar(stat = "identity",fill="#F9A68A",color = "#F9A68A", width = 0.8) +
      theme_classic()+
      ylab("-log10(Pvalue)") +
      xlab("")+
      xlim(0,300)+
      theme(axis.text.x = element_text(size = 15),
            axis.text.y = element_text(size = 15),
            axis.text = element_text(size = 15,colour = "black"),
            axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
            axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
            axis.title.x = element_text(size = 15,colour = "black"))
# ggsave(paste0(fig_dir,"lolipop_scstage_dmr_20240722.pdf"),plot = p4,width = 7,height = 4,dpi = 300)
print(p4)
```
## Figure S9 S10 S11 The same analysis process is shown in Figure 3E 3F

## Figure S12 C-F The same analysis process is shown in Figure 4A
## Figure S14  The same analysis process is shown in Figure 3 G-H
## Figure S15 The same analysis process is shown in Figure 4A

## Figure S16 2025-07-23 the lasted modified 

```{r}
## load data 
work_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/data/figS5-6_dat/"
meta_scstage_df <- readRDS("~/methylation/project/EM_seq/data/metadata/meta_scstage_df.rds")
scstage_go_element <-read.csv(paste0(work_dir,"scstage_element_go4_meth.csv"))
scstage_speci_go_dup_genes <- read.csv(paste0(work_dir,"scstage_speci4_go_dup_gene_meth.csv"))

## 
meta_scstage_df <- meta_scstage_df %>% mutate(sample_names=str_replace_all(sample,"GC","sc"))
meta_scstage_df$sample_list <- meta_scstage_df$sample_names
meta_scstage_df <- meta_scstage_df %>% mutate(sample_names=str_replace_all(sample_names,"-","."))
###

## check the duplication gene 
scstage_speci_go_dup_genes$term_gene <-  paste0(scstage_speci_go_dup_genes$Term,"_",scstage_speci_go_dup_genes$Genes)

## merge dup_term_gene to the orignal gene dataframe 
merged_df <- merge(scstage_go_element, scstage_speci_go_dup_genes, by = "Genes", all.x = TRUE)

merged_df_filter <- merged_df[,c(1,2,10,11,12,36,44)]

merged_df_filter$new_col <- apply(merged_df_filter[, c(3,4,5,1)], 1, function(x) paste(x, collapse = "_"))


merged_df_filter <- merged_df_filter %>%
  mutate(term_gene = if_else(is.na(term_gene) | term_gene == "NA", Term.x, term_gene))

library(dplyr)
df_new <- merged_df_filter  %>%
  group_by(new_col, Genes) %>% 
  summarise(go_terms = paste(unique(term_gene), collapse = "; "), .groups = "drop")

print(df_new)
df_new$Genes <- factor(df_new$Genes, levels = order_name)
sorted_df_new <- df_new %>% arrange(Genes)
```

```{r}
## 
scstage_go_element[,c(10,11,12,9)] <- lapply(scstage_go_element[,c(10,11,12,9)], as.character)
scstage_go_element$new_col <- apply(scstage_go_element[, c(10,11,12,9)], 1, function(x) paste(x, collapse = "_"))
h_dat <- scstage_go_element[scstage_go_element$p_adjust<0.05,]
```

```{r,fig.height=20,fig.width=20}

order_name <- c("CDH11","CDH13",##in 4 GO term
                "CDH10", "CDH18", "CDH2", "CDH26","CDH4", "CDH6" ,"CDH7" ,"CDH8", "CDH9",## 3GO term
                "PARD6B","TJP1",#3GO term
               "ADGRL3", "DSCAM", "PCDHB14","SDK2", "SLITRK1",##2GO term
               "CLDN11",
              "CTNND2","PKP2",
               "PRKCA",
               "PTPRT",## GO dup gene
              "TENM2","TENM3","EMB","LRRC4C","GRID2","CADM2",
              "CNTN6","UNC5D","IGSF11","PTPRD","LRFN5","CNTN4","NECTIN1",
              "SLITRK6","WDR1","GJA1","CD9","PTPRU","CLIC4","PTPRJ","MYLK","IGF1R",  
              "DACH1","DUSP10","FGF9","ARID2","PDGFRA","EDN1","ANXA1","HGF","MITF","TIAM1","TMEFF2", 
              "SFRP1","RAP2B","SPATA13","SINHCAF","SGK1","KANK1","LEF1","CXCR4","EGFR","SDCBP", 
              "ERBB4","PDGFD","MGAT5","PDGFC","FYN","CTNNA2","PLXNA4","TCAF2","XBP1","HACE1",  
             "YTHDF3","SULF1","TBX5","CXCL12","STK24","SNAI2","SPRY2","LGR6")

h_dat$Genes <- factor(h_dat$Genes, levels = order_name)
sorted_h_dat <- h_dat %>% arrange(Genes)

## add region elemnet information 
sorted_h_dat$elemnets[is.na(sorted_h_dat$elemnets)] <- "None"

sorted_go_dat <- sorted_h_dat[,c(1,9,14:36)]
sorted_h_dat[27,36] <- "chr4_61983982_61984475_ADGRL3_1"
sorted_h_dat[31,36] <- "chr17_73698076_73698175_SDK2_1"
sorted_h_dat[100,36] <- "chr8_58534885_58535347_SDCBP_1"

rownames(sorted_go_dat) <- sorted_h_dat$new_col
heatmap_dat <- sorted_go_dat[,c(3:23)]%>%as.matrix()

## pull sample names according to gender
sc_male <-meta_scstage_df %>% 
  filter(gender=="Male")%>%
  pull(sample_names)

sc_female <-meta_scstage_df %>% 
  filter(gender=="Female")%>%
  pull(sample_names)
heat_male_dat <- heatmap_dat[,sc_male,drop=FALSE]
heat_female_dat <- heatmap_dat[,sc_female,drop=FALSE]
sc_male_df <- meta_scstage_df[meta_scstage_df$sample_names%in%sc_male,]
sc_female_df <- meta_scstage_df[meta_scstage_df$sample_names%in%sc_female,]

status_colors <- c("Control" = "#63A6D0", "Early Stage" = "#F9A68A", "Advanced Stage" = "#D34447")
gender_colors <- c("Male" = "#AACBFE", "Female" = "#DE8686")
my_color=c("#4675FE","#F9F5C9","#FCA605")
ann_sc_male <- HeatmapAnnotation(Status=sc_male_df$stage,
                                  Gender=sc_male_df$gender,
                                  annotation_name_side = "left",
                                 annotation_name_gp = gpar(fontsize = 0),
                                  col=list(
                                  Status = status_colors,
                                  Gender = gender_colors))
ann_sc_female <- HeatmapAnnotation(Status=sc_female_df$stage,
                                  Gender=sc_female_df$gender,
                                  annotation_name_side = "right",
                                  col=list(
                                  Status = status_colors,
                                  Gender = gender_colors))
sorted_h_dat$elemnets <- as.factor(sorted_h_dat$elemnets)
## Save the output data
sorted_h_dat_output <- sorted_h_dat[,c(1,9:12,14:35)]
sorted_h_dat_output$chr_region <- paste0(sorted_h_dat_output$Chr,"_",sorted_h_dat_output$Start,"_",sorted_h_dat_output$End)

sorted_h_dat_output <- sorted_h_dat_output[,c(28,1:2,27,6:26)]
# write.csv(sorted_h_dat_output,paste0(fig_data,"figS11_gene_region.csv"))

row_enhancer <- c("None"="#F2F2F2","CTCF_binding_site"="#8DD3C7","enhancer"="#BEBADA",
                  "open_chromatin_region"="#FB8072","promoter"="#80B1D3","TF_binding_site"="#CCEBC5")
row_ann_element <- rowAnnotation(Elements=sorted_h_dat$elemnets,col=list(Elements=row_enhancer))
row_groups <- c(rep("GO:0098742;GO:0034329;GO:0007043;GO:0030334", 3),
                rep("GO:0098742;GO:0034329;GO:0007043",18),
                rep("GO:0034329;GO:0007043;GO:0030334",3),
                rep("GO:0098742;GO:0034329",9),
                rep("GO:0098742;GO:0007043",5),
                rep("GO:0007043;GO:0030334",1),
                rep("GO:0098742;GO:0030334",2),
                rep("GO:0098742",26),
                rep("GO:0034329",1),
                rep("GO:0007043",3),
                rep("GO:0030334",53))
row_annot <- rowAnnotation(GO = row_groups,
                           col = list(
                           GO=c("GO:0034329"="#F9F5C9",
                                "GO:0030334"= "#63A6D0",
                                "GO:0007043"="#4675FE",
                                "GO:0098742"="#DE8686",
                                "GO:0007043;GO:0030334"="#6C040F",
                                "GO:0034329;GO:0007043;GO:0030334"="#FCBC85",
                                "GO:0098742;GO:0034329"="#F96949",
                                "GO:0098742;GO:0007043" ="#FCEBDE",
                                "GO:0098742;GO:0030334"="#AACBFE",
                                "GO:0098742;GO:0034329;GO:0007043" = "#FC8B39",
                                "GO:0098742;GO:0034329;GO:0007043;GO:0030334" = "#D94500")),                         
                                annotation_legend_param = list(
                                  GO = list(
                                    at=c(
                                    "GO:0034329",
                                    "GO:0030334",
                                    "GO:0007043",
                                    "GO:0007043;GO:0030334",
                                    "GO:0098742",
                                    "GO:0034329;GO:0007043;GO:0030334",
                                    "GO:0098742;GO:0034329",
                                    "GO:0098742;GO:0007043",
                                    "GO:0098742;GO:0030334",
                                    "GO:0098742;GO:0034329;GO:0007043",
                                    "GO:0098742;GO:0034329;GO:0007043;GO:0030334"
                                  ),
labels=c("Cell Junction Assembly",
         "Regulation Of Cell Migration",
         "Cell-Cell Junction Assembly",
         "Cell-Cell Junction Assembly;Regulation Of Cell Migration",
         "Cell-Cell Adhesion Via Plasma-Membrane Adhesion Molecules",
         "Cell Junction Assembly;Cell-Cell Junction Assembly;Regulation Of Cell Migration",
         "Cell-Cell Adhesion Via Plasma-Membrane Adhesion Molecules;Cell Junction Assembly",
         "Cell-Cell Adhesion Via Plasma-Membrane Adhesion Molecules;Cell-Cell Junction Assembly",
         "Cell-Cell Adhesion Via Plasma-Membrane Adhesion Molecules;Regulation Of Cell Migration",
         "Cell-Cell Adhesion Via Plasma-Membrane Adhesion Molecules;Cell Junction Assembly;Cell-Cell Junction Assembly",
"Cell Junction Assembly;Cell-Cell Junction Assembly;Regulation Of Cell Migration;Cell-Cell Adhesion Via Plasma-Membrane Adhesion Molecules")
)
))
###
heat_map_1 <- Heatmap(heat_male_dat,
                      cluster_rows = T,
                      cluster_columns = T,
                      row_split = row_groups,
                      # right_annotation = c(row_annot,row_ann_element),
                      top_annotation =ann_sc_male ,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                    col = colorRampPalette(my_color)(100),
                    na_col = "white",
                    show_heatmap_legend = FALSE,
                     border_gp = gpar(col = "NA"),
                    width =unit(20,"cm"),
                     height = unit(40,"cm")
                      )

heat_map_2 <- Heatmap(heat_female_dat,
                      cluster_rows = T,
                      cluster_columns = T,
                      row_split = row_groups,
                      right_annotation = c(row_annot,row_ann_element),
                      top_annotation = ann_sc_female,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                    col = colorRampPalette(my_color)(100),
                    na_col = "white",
                    width =unit(10,"cm"),
                     height = unit(40,"cm")
                      )

p <- heat_map_1+heat_map_2
p
# pdf(paste0(fig_dir,"figS11_scstage_topGO_go_genes_20250325.pdf"),width = 25,height = 35)
# draw(p, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
# 
# dev.off()

```



## FigS17 A-B Make violin plot of scstage_dmr associated gene

```{r,echo=TRUE}
## import data

gene_dat <- "~/methylation/project/EM_seq/analysis/DMR_analysis/result/scstage_change_ann/"
fig_dir <- "~/methylation/project/EM_seq/TEMPT-paper-codes/DMR_meth_scstage_fig/"
smp_scsatge_list <- meta_scstage_df$sample_list

smp_scstage_DMR_meth <- read.table(paste0(sc_dat_dir,"scstage_dmr_change_region.sorted.txt"),col.names = "chr_region")
smp_scstage_DMR_cov <- read.table(paste0(sc_dat_dir,"scstage_dmr_change_region.sorted.txt"),col.names = "chr_region")

for (i_sample in smp_scsatge_list){
  a <- read.table(paste0(sc_dat_dir,i_sample,"_sc_dmr.calc.bed"),col.name = c("chromosome","start","end","mod","unmod","meth_level"))

 smp_scstage_DMR_meth[[i_sample]] <- a$meth_level
 smp_scstage_DMR_cov[[i_sample]] <- a$mod+a$unmod
}
## scstage_meth_dmr 2414 dmr_region
scstage_meth_dmr <- read.csv(paste0(gene_dat,"scstage_meth_data.csv"))
scstage_meth_dmr$p_adjust <- as.numeric(scstage_meth_dmr$p_adjust)
scstage_dat_ann <- scstage_meth_dmr

## trans data format
scstage_dat_ann[,c(1:5)] <- lapply(scstage_dat_ann[,c(1:5)], as.character)
scstage_dat_ann$chr_region <- apply(scstage_dat_ann[,c(1:3)], 1, function(x) paste(x,collapse = "_"))
scstage_dat_ann$gene_region <-  apply(scstage_dat_ann[, c(1:5)], 1, function(x) paste(x, collapse = "_"))
## define scstage_dmr gene annotated
scstage_dmr_ann <- scstage_dat_ann[,c(27,28,6:26)]

## merge data
merge_scstage_meth_ann <- merge(scstage_dmr_ann,smp_scstage_DMR_meth,by="chr_region")
scstage_meth_ann <- merge_scstage_meth_ann[,c(2,24:44)]

merge_scstage_cov_ann <- merge(scstage_dmr_ann,smp_scstage_DMR_cov,by="chr_region")
scstage_cov_ann <- merge_scstage_cov_ann[,c(2,24:44)]

## sc stage methylation level matrix 
data <- scstage_meth_ann
rownames(data) <- as.character(data$gene_region)
data <- data[, -1]
tr_data <- t(data)
rownames(tr_data) <- colnames(data)
colnames(tr_data) <- rownames(data)
tr_data_meth <- as.data.frame(tr_data)

## sc stage coverage in DMRs 
data_cov <- scstage_cov_ann
rownames(data_cov) <- as.character(data_cov$gene_region)
data_cov <- data_cov[, -1]
tr_data_co <- t(data_cov)
rownames(tr_data_co) <- colnames(data_cov)
colnames(tr_data_co) <- rownames(data_cov)
tr_data_cov <- as.data.frame(tr_data_co)
##

common_cols <- intersect(names(tr_data_cov), names(tr_data_meth))
count=0

p_value <- list()
## 
for (col in common_cols) {
  df <- data.frame(x = tr_data_cov[[col]], y = tr_data_meth[[col]]) 
  df$status <- meta_scstage_df$stage
  df$smp <- str_sub(rownames(tr_data_cov),1,4)
  df <- na.omit(df)
  ## do scatter plot 
    wilcox_test <- wilcox.test(df$y[df$status=="Early Stage"],df$y[df$status=="Advanced Stage"],paired = FALSE)
    p.value <- wilcox_test$p.value
    p_value[[col]] <- p.value
  if(!is.na(p.value)&&p.value< 0.05&&length(p.value)!=0){ 
        p1 <- ggplot(df, aes(x = status, y = y)) +
              geom_violin(aes(fill=status,alpha=1),scale="width") +
              scale_fill_manual(values =c("#D34447", "#F9A68A"))+
              geom_boxplot(aes(fill=status),
                            outlier.shape = NA,lwd=0.4,width=0.05)+
              scale_fill_manual(values =c("#FB4F4F", "#FE7E2A"),guide = 'none')+
                geom_jitter(aes(fill = status),
                          position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.8),
                          size = 3, shape = 21,)+
              labs(y ="Methylation level", x = paste("y_", col), title = paste0("p = ", round(p.value, 10)))+
              theme_classic()+
              theme(axis.text.x = element_text(size = 15),
                    axis.text.y = element_text(size = 15),
                    axis.text = element_text(size = 15,colour = "black"),
                    axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
                    axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
                    axis.title = element_text(size = 15,colour = "black"),
                    legend.position="none")
        # ggsave(plot = p1,
        #           filename =paste0(fig_dir,"meth_",col,"_violin_plot.pdf"),
        #        width = 6,height = 4,dpi = 300)
        print(p1)
         }else{
             print(paste0(col,"_wilcox p.adjust less than 0.05,and the p is ",p.value))
            count=count+1 }
          print(count)
 }

```

## Figure S18 A-B 

```{r}
# Tissue contribution ----
filename <- "~/methylation/project/EM_seq/analysis/deconv_analysis/TAPS_data/data/sample_mixtures.csv"
dat <- read.table(filename, header=TRUE, stringsAsFactors=FALSE, sep="\t")
fig_dir <- "~/methylation/project/EM_seq/analysis/deconv_analysis/TAPS_data/figure/"

dat.m <- melt(dat,id.vars=c("atlas_tissues"))
dat.m$atlas_tissues[dat.m$atlas_tissues %in% c("immature neutrophil","mature neutrophil")] <- "neutrophil"
dat.m$atlas_tissues[dat.m$atlas_tissues %in% c("immature conventional dendritic cell","mature conventional dendritic cell")] <- "dendritic cell"
a <- dat.m
dat.m <- aggregate(dat.m$value, by=list(Category=dat.m$variable, dat.m$atlas_tissues ), FUN = sum) %>% as.data.frame()
colnames(dat.m) <- c("variable","atlas_tissues","value")
dat.m$Type <- substr(dat.m$variable,1,2)
dat.m$Type <- recode(dat.m$Type,"sc"="Cancer","gp"="Benign")

tissue_group <- aggregate(dat.m$value, by=list(Category=dat.m$Type,dat.m$atlas_tissues ), FUN = mean) %>% as.data.frame()
colnames(tissue_group) <- c("status","tissue","value")

tissue_contribution_pie_plot <- function(type,clor){
  
  seldat <- tissue_group[tissue_group$status== type,]
  seldat$tissue[seldat$value<=0.015] <- "Other"
  seldat <- aggregate(seldat$value, by=list(Category=seldat$status,seldat$tissue ), FUN = sum) %>% as.data.frame()
  colnames(seldat) <- c("status","tissue","value")
  seldat$value <- round(seldat$value,2) * 100
  seldat <- seldat[order(seldat$value),]
  seldat$tissue <- factor(seldat$tissue,levels=seldat$tissue)
  seldat <- seldat %>%
    arrange(desc(tissue)) %>%
    mutate(lab.ypos = cumsum(value) - 0.5*value)
  nb.cols <- nrow(seldat)
if (clor=="Cancer") {
    ## define GC tissues and paired color 
    mycolors <- c("neutrophil"="#8DD3C7","spleen"="#FFFFB3","dendritic cell"="#BEBADA",
                
                "Thymus"="#FB8072","macrophage"="#80B1D3","monocyte"="#FDB462",
                
               "osteoclast"="#B3DE69","plasma_cell"="#FCCDE5","liver"="#66C2A5",
               
                "intestine"="#FFD92F","Other"="#B3B3B3")
  
  }else if (clor=="Benign") {
     ## define GP tissues and paired color 
 mycolors <- c("neutrophil"="#8DD3C7","spleen"="#FFFFB3","dendritic cell"="#BEBADA","monocyte"="#FDB462",
                
                "Thymus"="#FB8072","Other"="#B3B3B3","osteoclast"="#B3DE69", "macrophage"="#80B1D3",
               
                
               "naive_tcell"= "#E78AC3" ,"lung"="#4DAF4A",
               
                "heart"="#F0027F" )
  }else{
        stop("Error: Typo! GC or GP label expected!")
}    

  p <- ggplot(data = seldat, aes(x = "", y = value, fill = tissue)) + 
          geom_bar(width = 1, stat = "identity", color = "white") +
          coord_polar("y", start = 0)+
          geom_text(aes(y = lab.ypos, label = value), color = "white",size=3) +
          scale_fill_manual(values = mycolors) +
          theme_void() +
          ggtitle(type)

}

 p1 <- tissue_contribution_pie_plot("Cancer","Cancer")
 
 print(p1)
 p2 <- tissue_contribution_pie_plot("Benign","Benign")
 print(p2)
```

## Figure S18 C-D

```{r}
## load data 
meta_df <- readRDS("~/methylation/project/EM_seq/data/metadata/meta_df.rds")
sample_list <- meta_df$sample_list
dat_dir <- "~/methylation/project/EM_seq/analysis/deconv_analysis/TAPS_data/data/"
fig_dir <- "~/methylation/project/EM_seq/TEMPT-paper-codes/Too_FIGS_20250722/"

sample_mixtures <- read.csv(paste0(dat_dir,"sample_mixtures.csv"),sep = "\t")
df_1 <- sample_mixtures
col_value <- colnames(df_1)[2:39]

long_df_1 <- df_1%>%pivot_longer(cols =col_value,names_to = "sample",values_to = "probility")
long_df_1$smp_type <- substr(long_df_1$sample,1,2)
long_df_1 <- long_df_1%>%mutate(smp_type=recode(smp_type, "sc" = "GC", "gp" = "GP"))

long_df_1$smp_type <- as.factor(long_df_1$smp_type)

plots <- list()
p_value <- list()
 #i=19
for (i in unique(long_df_1$atlas_tissues)){

df_1 <- long_df_1[long_df_1$atlas_tissues==i,]
wilcox_rls <- wilcox.test(df_1$probility[df_1$smp_type == "GC"], 
                          df_1$probility[df_1$smp_type == "GP"],exact = FALSE)
p.val <- wilcox_rls$p.value

mean_fold_change_GC <- mean(df_1$probility[df_1$smp_type == "GC"])/mean(df_1$probility[df_1$smp_type == "GP"])
mean_fold_change_GP <- mean(df_1$probility[df_1$smp_type == "GP"])/mean(df_1$probility[df_1$smp_type == "GC"])

print(paste0("Mean ",i," fold change GC/GP is ",round(mean_fold_change_GC,3)))
print(paste0("Mean ",i," fold change GP/GC is ",round(mean_fold_change_GP,3)))

##do violin plot 
plot <- ggplot(df_1, aes(x = smp_type, y = probility)) +
  geom_violin(aes(fill=smp_type,alpha=1),scale = "width") +
  scale_fill_manual(values =c("#FC9595", "#C4E3FF"))+
  geom_boxplot(#aes(fill=smp_type),
                width=0.15,outlier.shape = NA)+
  geom_jitter(position=position_jitter(0.1), aes(fill=smp_type,color=smp_type),size=2)+
  scale_color_manual(values=c("#FB8D8D","#C0E1FF"))+
  #scale_fill_manual(values =c("#FB4F4F", "#93CDFF"),guide = 'none')+
  labs(y =paste0("Probability of originating"), x =" ", 
       title = paste0("cell type ", i, "\n",
                      "p = ", round(p.val, 3) ,"\n",
                      "mean fold change GC/GP  ",round(mean_fold_change_GC,3),"\n",
                      "mean fold change GP/GC  ",round(mean_fold_change_GP,3)))+
  theme_classic()+ 
  theme(
     axis.text.x = element_text(size = 15),
         axis.text.y = element_text(size = 15),
         axis.text = element_text(size = 15,colour = "black"),
         axis.title = element_text(size = 15,colour = "black"),
        legend.position="none")
# ggsave(plot,filename=paste0(fig_dir,"cell_type_",i,"figS18c_d_TOO_20250723.pdf"),width = 8,height = 6,dpi = 300)

print(plot)
plots[[i]] <- plot
p_value[[i]] <- p.val
}
p_dataframe <- data.frame(tissue_type=names(p_value),
                             p_val=unlist(p_value))
p_dataframe$p_adjust <- p.adjust(p_dataframe$p_val,method = "BH")
 print(p_dataframe)

```
## Figure S19 The same analysis process is shown in Figure S17

## Technical 

```{r}
sessionInfo()
```


