---
title: "Figure 3"
author: "Zhenna Jiao"
date: "2024-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(VennDiagram)
library(ggplot2)
library(readxl)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
library(scales)
library(readr)
library(tidyverse)
library(reshape2)
library(dplyr)
library(caret)
library(glmnet)
library(ROCR)
library(pryr)
library(cowplot)
library(ggplotify)
library(RColorBrewer)
library(MLmetrics)
library(GenomicRanges)
library(plotROC)
library(data.table)
library(magrittr)
library(pROC)
library(randomForest)
library(gbm)
library(ggbiplot)
library(corrplot)
library(ggrepel)
```
## Import data

```{r cars}
dat_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/data/ref_dmr_ann_data/"
fig_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/figures/fig3.plot/"
meth_dmr <-  read.csv(paste0(dat_dir,"fil_dmr_meth_gene.csv"))
meta_df <- readRDS("~/methylation/project/EM_seq/data/metadata/meta_df.rds")
meta_df <- meta_df%>% mutate(sample_names=str_replace_all(sample_list,"-","."))
```
## Fig 3A Plot a heatmap for the correlation between GP and GC on DMRs regions

```{r,fig.width=14,fig.height=10}
## Import data
dmr_dat_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/data/filter_sample/dmr_file_delta0.15/"
sample_DMR_meth <- readRDS(paste0(dmr_dat_dir,"sample_DMR_meth.rds"))
sample_DMR_meth_nona <- na.omit(sample_DMR_meth)

## Perform correlation analysis without removing missing values (use the parameter "pairwise")
data_keep_na <-  sample_DMR_meth
row.names(data_keep_na) <- as.character(sample_DMR_meth$chr_region)
data_keep_na <- data_keep_na[,-1]
colnames(data_keep_na) <- meta_df$sc_smp
DMR_cor_keep_na <- cor(data_keep_na,use = "pairwise.complete.obs")

## Plot a heatmap
annotation_col_keepna <- data.frame(Type=meta_df$sc_type,Stage=meta_df$sc_stage,Gender=meta_df$sc_gender)
DMR_cor_df_keepna <- as.data.frame(DMR_cor_keep_na)
rownames(annotation_col_keepna)  <- colnames(DMR_cor_df_keepna)
my_color=c("#4675FE","white","#F9A68A","#D34447")
type_colors=list(Type=c(GP="#C0E1FF",GC="#FB8D8D"))
gender_color=list(Gender=c(Male="#AACBFE",Female="#DE8686"))
stage_color=list(Stage=c("Control"="#63A6D0","Early Stage"="#F9A68A","Advanced Stage"="#D34447"))

# pdf(paste0(fig_dir,"heatmap_smp_dmr_cor.pdf"),width = 14,height = 10)
pheatmap::pheatmap(DMR_cor_df_keepna,cluster_rows = T ,
                   cluster_cols =T,show_rownames = T,show_colnames = T,
                   annotation_col = annotation_col_keepna,
                   annotation_colors = c(type_colors,gender_color,stage_color),
                   scale = "none",
                   fontsize = 15,                
                  fontsize_row = 15,             
                  fontsize_col = 15,
                   cellwidth = 13,
                  cellheight =12,
                  color = colorRampPalette(my_color)(100))

# dev.off()
```
## Fig 3B  Calculate PCA for DMR between GC and GP based on the DMR region's methylation level (the matrix removes missing)

```{r,fig.width=10,fig.height=8}
# Import data
sample_DMR_meth_nona <- sample_DMR_meth%>%na.omit()
data <- sample_DMR_meth_nona
rownames(data) <- as.character(data$chr_region)
data <- data[, -1]
tr_data <- t(data)
rownames(tr_data) <- colnames(data)
colnames(tr_data) <- rownames(data)
tr_data <- as.data.frame(tr_data)
tr_data$sample_state <- meta_df$sc_type %>% as.factor()
tr_data$sample_stage <- meta_df$sc_stage  %>% as.factor()
# calculate PCA
tr.pca_all <- prcomp(tr_data[,1:28])
# do plot for pc1 and pc2
df1 <- tr.pca_all$x
pc_one <- as.data.frame(df1,tr_data$sample_state)
pc_one$Type <- tr_data$sample_state
pc_one$smp_name <- substr(rownames(tr_data),1,4)

summ<-summary(tr.pca_all)
xlab<-paste0("PC1(",round(summ$importance[2,1]*100,2),"%)")
ylab<-paste0("PC2(",round(summ$importance[2,2]*100,2),"%)")
p2<-ggplot(data = pc_one,aes(x=PC1,y=PC2,color=Type))+
  geom_point(size=5)+
  labs(x=xlab,y=ylab)+
  scale_color_manual(values = c("#FB8D8D", "#C0E1FF"))+
  guides(fill=F)+
  theme_classic()+
 theme(axis.text.x = element_text(size = 15),
              axis.text.y = element_text(size = 15),
              axis.text = element_text(size = 15,colour = "black"),
            axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
           axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
           axis.title = element_text(size = 15,colour = "black"))
 
 # ggsave(paste0(fig_dir,"pca_smp_dmr_removena.pdf"),plot = p2,width = 8,height = 5,dpi = 300)
print(p2)
```
## Fig 3C  Plot a lollipop for the annotated gene elements

```{r}
proj_dir <- "~/methylation/project/EM_seq/analysis/DMR_analysis/result/annotation_dmr/"
dat_smp <- read_excel(paste0(proj_dir,"dmr_features_20240722.xlsx"),sheet = "all_dmr_feature_fisher")
dat_smp$logp <- -log(dat_smp$two_tail,10)
## Plot a lollipop for DMR elements
p2 <- ggplot(dat_smp, aes(y = elments, x = logp)) +
      geom_bar(stat = "identity",fill="#63A6D0",color = "#63A6D0",width=0.8) +
      theme_classic()+
      ylab("-log10(Pvalue)") +
      xlab("")+
      theme(axis.text.x = element_text(size = 15),
            axis.text.y = element_text(size = 15),
            axis.text = element_text(size = 15,colour = "black"),
            axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
            axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
            axis.title.x = element_text(size = 15,colour = "black"))

# ggsave(paste0(fig_dir,"lolipop_smp_dmr_20240806.pdf"),plot = p2,width = 6,height = 4,dpi = 300)
print(p2)
```
## Fig 3D AUC score

```{r}
samples <-  meta_df$sample_list
meta <- as.data.frame(samples)
meta$status <- as.factor(c(rep("gp",17),rep("sc",21)))
meta$smp <- gsub("^(.{4}).*", "\\1", meta$samples)
set.seed(202404)
folds <- list()
# Randomly shuffle samples
fold_sample <- sample(meta$samples)
folds$fold1 <- fold_sample[1:8]
folds$fold2 <- fold_sample[9:16]
folds$fold3 <- fold_sample[17:24]
folds$fold4 <- fold_sample[25:31]
folds$fold5 <- fold_sample[32:38]
label="sc"

## Define Random Forest function
res_RF_regression_list <- list()
  DMRpredict_RF_regression <- function(mat,i){
        rownames(ratio_meth) <- as.character(ratio_meth$chr)
        ratio_meth <- ratio_meth[, -1]%>%t()
        ratio_meth <- as.data.frame(ratio_meth)
        label="sc"
        ratio_meth$smp_stage <- as.numeric(meta$status==label)
        all_sample <- meta$samples
        intrain_smp <- c(meta$samples[!meta$samples %in% folds[[i]]])
        intest_smp <- c(folds[[i]])
       
        real_label_train <- as.numeric(meta$status[meta$samples %in% intrain_smp]==label) %>% as.factor()
        real_label_test  <- as.numeric(meta$status[meta$samples %in% intest_smp]==label)  %>% as.factor()
        ## split train set and test set
        intrain_id <- rownames(ratio_meth) %in% intrain_smp
        intest_id <- rownames(ratio_meth) %in% intest_smp
        ##fit model
        set.seed(20240705)
        RF_ob <- randomForest(smp_stage ~ .,
                      data = ratio_meth,
                      subset = intrain_id,
                      importance=TRUE,proximity=TRUE,
                      maxnodes=10)
        
        varImpPlot(RF_ob)
        pred_train1 <- predict(RF_ob)
        pred_test1  <- predict(RF_ob,newdata = ratio_meth[intest_id,],type = "response" ) 
        confusion_matrix <- table(pred_test1,real_label_test)
        AUC_train1 <- roc(as.ordered(real_label_train),as.ordered(pred_train1))
        AUC_test2 <- roc(as.ordered(real_label_test),as.ordered(pred_test1)) 
        cat("this is RF round ",i,"pred_test_cvglm = ",names(pred_test1),"\n")
        cat("this is RF round ",i,"pred_test_cvglm = ",pred_test1,"\n")
        cat("this is RF round ",i,"auc_cvglm_train = ")
        print(AUC_train1)
        cat("this is RF round ",i,"auc_cvglm_test = ")
        print(AUC_test2)
        res_RF_regession <- list("smp" = intest_smp,
                    "pred_test_model" = pred_test1,
                    "auc_RF_train" = AUC_train1,
                    "auc_RF_test" = AUC_test2 ,
                    "fit_model" = RF_ob,
                    "confusion_matrix"=confusion_matrix)
        return(res_RF_regession)
        cat("complete round",i ,"\n")
 }
 
## Define a function
 
 pred_dat_list <- list()
 DOPLOT <- function(folds,list,all_sample,label){
  
  for (i in 1:length(folds)){
    pred_dat_list[[i]] <- do.call(cbind, lapply(list[i], function(x) x$pred_test_model))
}

pred_dat <- do.call(rbind, pred_dat_list)
pred_dat <- as.data.frame(pred_dat)
colnames(pred_dat) <- "pred_test_cvglm"
pred_dat$status <- substr(row.names(pred_dat),1,2)
pred_dat$samples <- row.names(pred_dat)
pred_dat$label <- as.numeric(pred_dat$status[pred_dat$samples %in% all_sample]==label)
pred_dat$smp_name <- substr(pred_dat$samples,1,4)
pred <- prediction(pred_dat$pred_test_cvglm, pred_dat$label)
perf_auc <- performance(pred, "auc")
perf_roc <- performance(pred, measure = "tpr", x.measure = "fpr")
plot( perf_roc)

### Plot 
p1 <- tibble(x=perf_roc@x.values[[1]], y=perf_roc@y.values[[1]]) %>%
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color=cut(y, c(-1,0.5,0.7,0.9,1)))) +
  geom_point(aes(color=cut(y, c(-1,0.5,0.7,0.9,1)))) +
  theme_bw()+
  labs(color="TPR", y="True positive rate", x="False positive rate") +
  scale_color_brewer(palette = "Reds") +
  theme(legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.text = element_text(size = 12),
  axis.title = element_text(face = "bold"))+
  geom_text(aes(x=1, y=0, label=paste0("AUC: ",perf_auc@y.values[[1]]%>%round(2))),
            hjust="right", vjust="bottom")
print(p1)

  fac <- with(pred_dat, reorder(smp_name, pred_test_cvglm, median, order = TRUE))
  pred_dat$smp_name <- factor(pred_dat$smp_name, levels = levels(fac))
  ## Replace "sc" to "GC, "gp" to "GP" using the mutate function
  dat <- pred_dat
  dat <- dat%>%mutate(smp=smp_name%>%gsub("^sc","GC",.)%>%gsub("^gp","GP",.))
  fac_1 <- with(dat, reorder(smp, pred_test_cvglm, median, order = TRUE))
  dat$smp <- factor(dat$smp, levels = levels(fac_1))
  dat$Type <- substr(dat$smp,1,2)
  p2 <- ggplot(dat, aes(x=smp, y=pred_test_cvglm ,color=Type)) +
        geom_point(size=3) +
        theme_classic()+
        ylab("GC cancer score")+
        scale_color_manual(values = c("#FB8D8D", "#C0E1FF"))+
         theme(axis.text.x = element_text(size = 12,angle = 45,vjust = 0.5),
          axis.text.y = element_text(size = 15),
          axis.text = element_text(size = 15,colour = "black"),
          axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
          axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
          axis.title = element_text(size = 15,colour = "black"))
  print(p2)
  
   p3 <- ggplot() +
    geom_roc(data=pred_dat, aes(d = pred_dat$label, m = pred_test_cvglm), labels = F, color="black") +
    theme_classic() +
    annotate(geom="text", x=0.8, y=0.6, label=paste0(" AUC: ",perf_auc@y.values[[1]]%>%round(2)),color="black",size=5) +
    ggtitle("AUC")+
    labs(color="TPR", y="True positive rate", x="False positive rate")+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(), axis.line = element_line(colour = "black"),
          axis.text.x = element_text(size = 15),
          axis.text.y = element_text(size = 15),
          axis.text = element_text(size = 15,colour = "black"),
          axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
          axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
          axis.title = element_text(size = 15,colour = "black"))
  ggsave(plot = p3,filename = paste0(fig_dir,"RandomForest_ROCplot_20240827.pdf"),
        width = 8,height = 6,dpi = 300)
  print(p3)

 }

## Define a function to plot each fold ROC
 all_roc_data <- data.frame()
auc_labels <- data.frame()
PlotForAllFold <- function(list,folds,all_sample,label){
for (i in 1:length(folds)){
    pred_dat <- list[[i]][["pred_test_model"]]%>%as.data.frame()
    colnames(pred_dat) <- "pred_test_cvglm"
    pred_dat$status <- substr(row.names(pred_dat),1,2)
    pred_dat$samples <- row.names(pred_dat)
    pred_dat$label <- as.numeric(pred_dat$status[pred_dat$samples %in% all_sample]==label)
    pred_dat$smp_name <- substr(pred_dat$samples,1,4)
    ### calculate AUC score for fold i
    pred <- prediction(pred_dat$pred_test_cvglm, pred_dat$label)
    perf_auc <- performance(pred, "auc")
    perf_roc <- performance(pred, measure = "tpr", x.measure = "fpr")
    ### do plot
    plot(perf_roc, col = "blue", main = "ROC Curve", lwd = 2)
    abline(0, 1, col = "gray", lty = 2, lwd = 1)
    text(0.5, 0.3, paste("AUC =", round(perf_auc@y.values[[1]], 2)), adj = c(0.5, 0.5), col = "black", cex = 1.5)
        roc_data <- data.frame(
        FPR = perf_roc@x.values[[1]],
        TPR = perf_roc@y.values[[1]],
        Fold = paste("Fold", i)
    )
    all_roc_data <- rbind(all_roc_data, roc_data)
    print(paste("Fold", i, "AUC:", round(perf_auc@y.values[[1]], 2)))
    auc_labels <- rbind(auc_labels, data.frame(
        Fold = paste("Fold", i),
        AUC = round(perf_auc@y.values[[1]], 2),
        x = 0.8,
        y = 0.5 - (i * 0.05)
    ))
}
  colors <- brewer.pal(n = length(folds), name = "Set1")

 p <-  ggplot(all_roc_data, aes(x = FPR, y = TPR, color = Fold)) +
    geom_line(size = 1) +
    theme_classic() +
    ggtitle("ROC Curves for All Folds") +
    xlab("False Positive Rate") +
    ylab("True Positive Rate") +
    scale_color_manual(values = colors) +
    theme(legend.title = element_blank()) +
    xlim(0,1)+
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray")+
    geom_text(data = auc_labels, aes(x = x, y = y, label = paste(Fold, "AUC:", AUC)),
              color = "black", size = 4, hjust = 0)+
    theme(axis.text.x = element_text(size = 15),
          axis.text.y = element_text(size = 15),
          axis.text = element_text(size = 15,colour = "black"),
          axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
          axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
          axis.title = element_text(size = 15,colour = "black"))
    
    # ggsave(plot = p,filename = paste0(fig_dir,"RandomForest_eachfold_ROCplot.pdf"),
    #      width = 8,height = 6,dpi = 300)
print(p)

}

## Import data 
dmr_fold_proj <- "~/methylation/project/EM_seq/analysis/DMR_analysis/result/fit_model_dmr/"
for (i in 1:length(folds)){
     dmr_meth_matrix <- read.table(paste0(dmr_fold_proj,"dss_result_fold_",i,"/dmr_fold_",i,"_region.sorted.filter.txt"),col.names = "chr")
     dmr_cov_matrix <- read.table(paste0(dmr_fold_proj,"dss_result_fold_",i,"/dmr_fold_",i,"_region.sorted.filter.txt"),col.names = "chr")
   for (i_sample in samples){
      a <- read.table(paste0(dmr_fold_proj,"dss_result_fold_",i,"/sample_dmr_filter_file/",i_sample,
                             "_fold_",i,"_dmr.calc.bed"),
                      col.name = c("chromosome","start","end","mod","unmod","meth_level"))
      dmr_meth_matrix[[i_sample]] <- a$meth_level
      dmr_cov_matrix[[i_sample]] <- a$mod+a$unmod
   }
      dmr_meth <- na.omit(dmr_meth_matrix)
      dmr_cov <- na.omit(dmr_cov_matrix)
      ## filter by coverage threshold value is >=5,>=10,>=15,>=20 to get the sample meth_matrix 
      rownames(dmr_cov) <- dmr_cov$chr
      rownames(dmr_meth) <- dmr_meth$chr
      val=15
      data <- dmr_cov
      data1 <- dmr_cov[rowSums(data[,2:39] > val ) == 38 , ]
      cat("this is fold:",i,"-the dmr region left :",nrow(data1),"\n")
      meth_matrix <- merge(dmr_meth,data1,by="chr",suffixes=c(" "," "))
      ratio_meth <- meth_matrix[,1:(length(samples) + 1)]
      colnames(ratio_meth) <- colnames(dmr_meth)
      res_RF_regression_list[[i]] <- DMRpredict_RF_regression(mat = ratio_meth,i=i)
      #saveRDS(dmr_meth_matrix,paste0(dmr_fold_proj,"dss_result_fold_",i,"/dmr_fold_",i,"_meth_matrix.rds"))
}

DOPLOT(folds = folds,
       list = res_RF_regression_list,
       all_sample = meta$samples,
       label = "sc")
PlotForAllFold(list =res_RF_regression_list,folds = folds,all_sample = meta$samples,label="sc")
```
## Fig 3E AUC score for the independent validation cohort 

```{r}
## import data 
dat_met <- "~/methylation/project/EM_seq/20241125_data_reanalysis/metadata/sample_metadata_20250207.xlsx"
meta <- read_excel(dat_met, sheet = "filter_added_sample")
meta$Sample_name <- paste0(meta$Sample,"_",meta$Sample_ID)
meta$type <- substr(meta$Sample,1,2)
meta$sc_type <-  toupper(substr(meta$Sample,1,2))
meta$Sample_upper <- toupper(meta$Sample)
meta$sample_id <- paste0(meta$Sample,"_",meta$Sample_ID)
## Partition the samples into nine folds randomly
set.seed(20250226)
gc_samples <- meta$Sample[meta$type == "gc"]
gp_samples <- meta$Sample[meta$type == "gp"]

gc_samples <- sample(gc_samples)
gp_samples <- sample(gp_samples)
folds <- list()
folds$fold1 <- c(gc_samples[1], gp_samples[1])
folds$fold2 <- c(gc_samples[2], gp_samples[2])
folds$fold3 <- c(gc_samples[3], gp_samples[3])
folds$fold4 <- c(gc_samples[4], gp_samples[4])
folds$fold5 <- c(gc_samples[5], gp_samples[5])
folds$fold6 <- c(gc_samples[6], gp_samples[6])
folds$fold7 <- c(gc_samples[7], gp_samples[7])
folds$fold8 <- c(gc_samples[8], gc_samples[9], gp_samples[8])
folds$fold9 <- c(gc_samples[10], gc_samples[11], gp_samples[9])
print(folds)
## Define Random Forest function
res_RF_regression_list <- list()
DMRpredict_RF_regression <- function(mat,i,label=NULL,meta,train_mat){
        rownames(ratio_meth) <- as.character(ratio_meth$chr)
        ratio_meth <- ratio_meth[, -1]%>%t()
        ratio_meth <- as.data.frame(ratio_meth)
        ratio_meth$smp_stage <- as.numeric(meta$sc_type==label)
        all_sample <- meta$Sample_ID
        intrain_smp <- c(meta$Sample_ID[!meta$Sample %in% folds[[i]]])
        intest_smp <- c(meta$Sample_ID[meta$Sample %in% folds[[i]]])
        real_label_train <- as.numeric(meta$sc_type[meta$Sample_ID %in% intrain_smp]==label) %>% as.factor()
        real_label_test  <- as.numeric(meta$sc_type[meta$Sample_ID %in% intest_smp]==label)  %>% as.factor()
        intrain_id <- rownames(ratio_meth) %in% intrain_smp
        intest_id <- rownames(ratio_meth) %in% intest_smp
        set.seed(20240705)
        RF_ob <- randomForest(smp_stage ~ .,
                      data = ratio_meth,
                      subset = intrain_id,
                      importance=TRUE,proximity=TRUE,
                      maxnodes=15)
        varImpPlot(RF_ob)
        pred_train1 <- predict(RF_ob)
        pred_test1  <- predict(RF_ob,newdata = ratio_meth[intest_id,])
        confusion_matrix <- table(pred_test1,real_label_test)
        
        AUC_train1 <- roc(as.ordered(real_label_train),as.ordered(pred_train1))
        AUC_test2 <- roc(as.ordered(real_label_test),as.ordered(pred_test1))
        
        cat("this is RF round ",i,"pred_test_cvglm = ",names(pred_test1),"\n")
        cat("this is RF round ",i,"pred_test_cvglm = ",pred_test1,"\n")
        cat("this is RF round ",i,"auc_cvglm_train = ")
        print(AUC_train1)
        cat("this is RF round ",i,"auc_cvglm_test = ")
        print(AUC_test2)
        
        res_RF_regession <- list("smp" = intest_smp,
                    "pred_test_model" = pred_test1,
                    
                    "auc_RF_train" = AUC_train1,
                    "auc_RF_test" = AUC_test2 ,
                    "fit_model" = RF_ob,
                    "confusion_matrix"=confusion_matrix)
        
        return(res_RF_regession)
        cat("complete round",i ,"\n")
}
## define plot function
pred_dat_list <- list()
DOPLOT <- function(folds,list,all_sample,label){
  for (i in 1:length(folds)){
    pred_dat_list[[i]] <- do.call(cbind, lapply(list[i], function(x) x$pred_test_model))
}

pred_dat <- do.call(rbind, pred_dat_list)
pred_dat <- as.data.frame(pred_dat)
colnames(pred_dat) <- "pred_test_cvglm"
pred_dat$status <- meta$sc_type[match(rownames(pred_dat),meta$Sample_ID)]
pred_dat$samples <- row.names(pred_dat)
pred_dat$label <- as.numeric(pred_dat$status[pred_dat$samples %in% meta$Sample_ID]==label)

pred_dat$smp_name <- meta$Sample[match(rownames(pred_dat),meta$Sample_ID)]%>%toupper()

pred <- prediction(pred_dat$pred_test_cvglm, pred_dat$label)
perf_auc <- performance(pred, "auc")
perf_roc <- performance(pred, measure = "tpr", x.measure = "fpr")
plot( perf_roc)

### plot 
p1 <- tibble(x=perf_roc@x.values[[1]], y=perf_roc@y.values[[1]]) %>%
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color=cut(y, c(-1,0.5,0.7,0.9,1)))) +
  geom_point(aes(color=cut(y, c(-1,0.5,0.7,0.9,1)))) +
  theme_bw()+
  labs(color="TPR", y="True positive rate", x="False positive rate") +
  scale_color_brewer(palette = "Reds") +
  theme(legend.position = "bottom",
  legend.title = element_text(face = "bold"),
  legend.text = element_text(size = 12),
  axis.title = element_text(face = "bold"))+
  geom_text(aes(x=1, y=0, label=paste0("AUC: ",perf_auc@y.values[[1]]%>%round(2))),
            hjust="right", vjust="bottom")
print(p1)

  fac <- with(pred_dat, reorder(smp_name, pred_test_cvglm, median, order = TRUE))
  pred_dat$smp_name <- factor(pred_dat$smp_name, levels = levels(fac))
  dat <- pred_dat
  p2 <- ggplot(dat, aes(x=smp_name, y=pred_test_cvglm ,color=status)) +
        geom_point(size=3) +
        theme_classic()+
        ylab("GC cancer score")+
        scale_color_manual(values = c("#FB8D8D", "#C0E1FF"))+
         theme(axis.text.x = element_text(size = 12,angle = 45,vjust = 0.5),
          axis.text.y = element_text(size = 15),
          axis.text = element_text(size = 15,colour = "black"),
          axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
          axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
          axis.title = element_text(size = 15,colour = "black"))
    
  # ggsave(paste0(fig_dir,"FOLD9_prediction_score.pdf"),plot = p2,width = 8,height = 6,dpi = 300)
print(p2)
  
p3 <- ggplot() +
    geom_roc(data=pred_dat, aes(d = pred_dat$label, m = pred_test_cvglm), labels = F, color="black") +
    theme_classic() +
    annotate(geom="text", x=0.8, y=0.6, label=paste0(" AUC: ",perf_auc@y.values[[1]]%>%round(2)),color="black",size=5) +
    ggtitle("AUC")+
    labs(color="TPR", y="True positive rate", x="False positive rate")+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(), axis.line = element_line(colour = "black"),
          axis.text.x = element_text(size = 15),
          axis.text.y = element_text(size = 15),
          axis.text = element_text(size = 15,colour = "black"),
          axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
          axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
          axis.title = element_text(size = 15,colour = "black"))
# ggsave(plot = p3,filename = paste0(fig_dir,"RandomForest_ROCplot_20250306.pdf"),
# width = 8,height = 6,dpi = 300)
print(p3)

 }

## Import data 
dmr_fold_proj <- "~/methylation/project/EM_seq/20241125_data_reanalysis/validation_for_the_added_samples/results/validation_filter_sample_9fold_CV/"
for (i in 1:9){
     dmr_meth_matrix <- read.table(paste0(dmr_fold_proj,"fold_",i,"/dmrs_call_fold_",i,
                                          "_DMR_region_sort_remove_blacklist_onecol.bed"),
                                   col.names = "chr")
     dmr_cov_matrix <- read.table(paste0(dmr_fold_proj,"fold_",i,"/dmrs_call_fold_",i,
                                          "_DMR_region_sort_remove_blacklist_onecol.bed"),
                                  col.names = "chr")
     
   for (i_sample in meta$Sample_ID){
      a <- read.table(paste0(dmr_fold_proj,"fold_",i,"/add_sample_dmr_bmeth_filter_file/",i_sample,"_dmr_meth.bed"),
                      col.name = c("chromosome","start","end","mod","unmod","meth_level"))
      dmr_meth_matrix[[i_sample]] <- a$meth_level
      dmr_cov_matrix[[i_sample]] <- a$mod+a$unmod
   }
     
      dmr_meth_train_set <- na.omit(dmr_meth_matrix)
      dmr_cov_train_set <-  na.omit(dmr_cov_matrix)
    
      rownames(dmr_meth_train_set) <- dmr_meth_train_set$chr
      rownames(dmr_cov_train_set) <- dmr_cov_train_set$chr
      
      val=9
      
      data <- dmr_cov_train_set
      data1 <- dmr_cov_train_set[rowSums(data[,2:ncol(dmr_cov_train_set)] > val ) 
                                  == length(meta$Sample_ID) , ]
      cat("this is fold:",i,"-the dmr region left :",nrow(data1),"\n")
      ## Define the trainning set meth matrix
      train_meth_matrix <- merge(dmr_meth_train_set,data1,by="chr",suffixes=c(" "," "))
      train_ratio_meth <- train_meth_matrix[,1:(length(meta$Sample_ID) + 1)]
      colnames(train_ratio_meth) <- colnames(dmr_meth_train_set)

      ratio_meth <- train_ratio_meth
      ## Save the filtered ratio_meth 
      # write.csv(ratio_meth,paste0(dmr_fold_proj,"fold_",i,"/fold_",i,"_predection_DMR_meth.csv"))
      
      res_RF_regression_list[[i]] <- DMRpredict_RF_regression(mat=ratio_meth,
                                                              i=i,label = "GC",meta = meta)
}

DOPLOT(folds = folds,
       list = res_RF_regression_list,
       all_sample = meta$Sample_ID,
       label = "GC")
```
## Fig 3F Plot a heatmap for the top 50 genes

```{r,fig.width=16,fig.height=10}
meth_dmr$p_adjust <- as.numeric(meth_dmr$p_adjust)
p_data <- meth_dmr[meth_dmr$p_adjust < 0.05,] %>% na.omit()
h_data <- p_data[,6:44]
rownames(h_data) <- p_data$Genes
h_data <- h_data[,-1]%>%as.matrix()

## Extract top 50 genes 
meth_dmr$p_adjust <- as.numeric(meth_dmr$p_adjust)
p_data <- meth_dmr[meth_dmr$p_adjust< 0.05,]

## Oreder the dataframe 
p_data <- p_data[order(p_data$p_adjust),]
h_data <- p_data[1:50,]
h_data[,c(6,2,3,4)] <- lapply(h_data[,c(6,2,3,4)],as.character)
h_data$new_col <- apply(h_data[,c(6,2,3,4)],1,function(x) paste(x,collapse = "_"))

## Extract specified rows to plot a heatmap
top20_gene_dat <- h_data[1:20,6:44]
top20_30_gene_dat <- h_data[21:50,6:44]
rownames(top20_gene_dat) <- top20_gene_dat$Genes
rownames(top20_30_gene_dat) <- top20_30_gene_dat$Genes
top20_gene_dat <- top20_gene_dat[,2:39]%>%as.matrix()
top20_30_gene_dat <- top20_30_gene_dat[,2:39]%>%as.matrix()

## Divided the matrix based on the gender 
male_samples <- meta_df %>%
  filter(sc_gender == "Male") %>%
  pull(sample_names)
female_samples <- meta_df %>%
  filter(sc_gender == "Female") %>%
  pull(sample_names)
top20_gene_dat_male <- top20_gene_dat[,male_samples,drop = FALSE]
top20_gene_dat_female <- top20_gene_dat[,female_samples,drop = FALSE]
top20_30_gene_dat_male <- top20_30_gene_dat[,male_samples,drop = FALSE ]
top20_30_gene_dat_female <- top20_30_gene_dat[,female_samples,drop = FALSE ]

## Add annotations to the heatmap
annotation_cols <-
  data.frame(Type=meta_df$sc_type,Stage=meta_df$sc_stage,Gender=meta_df$sc_gender,sample_names=meta_df$sample_names)
annotation_cols_male <- annotation_cols %>% filter(Gender == "Male")
annotation_cols_female <- annotation_cols %>% filter(Gender == "Female")
rownames(annotation_cols) <- colnames(top20_gene_dat)
rownames(annotation_cols_male) <- colnames(top20_gene_dat_male)
rownames(annotation_cols_female) <- colnames(top20_gene_dat_female)
type_colors <- c("GC" = "#FB8D8D", "GP" = "#C0E1FF")
status_colors <- c("Control" = "#63A6D0", "Early Stage" = "#F9A68A", "Advanced Stage" = "#D34447")
gender_colors <- c("Male" = "#AACBFE", "Female" = "#DE8686")
my_color=c("#4675FE","#F9F5C9","#FCA605")

male_df <- meta_df[meta_df$sample_names %in%male_samples,]
female_df <- meta_df[meta_df$sample_names %in%female_samples,]

##-------------- annotation for the heatmap--------------------------------
ann_male <- HeatmapAnnotation(
                            Status=male_df$sc_stage,
                             Gender=male_df$sc_gender,
                              Type=male_df$sc_type,
                            annotation_name_gp = gpar(fontsize = 0),
                               col=list(
                                 Type = type_colors,
                                  Status = status_colors,
                                  Gender = gender_colors))
ann_female <- HeatmapAnnotation(
                            Status=female_df$sc_stage,
                            Gender=female_df$sc_gender,
                            Type=female_df$sc_type,
                            annotation_name_gp = gpar(fontsize = 0),
                            col=list(Type = type_colors,
                                      Status = status_colors,
                                      Gender = gender_colors),
                                      annotation_name_side = "right")

ann_heatmap_df <- HeatmapAnnotation(
                            Status=meta_df$sc_stage,
                             Gender=meta_df$sc_gender,
                              Type=meta_df$sc_type,
                             annotation_name_side = "right",
                               col=list(
                                 Type = type_colors,
                                  Status = status_colors,
                                  Gender = gender_colors))


### ----------------heatmap plot-----------------------------------
heat_male <- Heatmap(top20_gene_dat_male,cluster_rows = T,
                      cluster_columns = T,
                      top_annotation = ann_male,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                      col = colorRampPalette(my_color)(100),
                      na_col = "white",width =unit(8,"cm"),
                     height = unit(10,"cm"),
                     show_heatmap_legend = FALSE,
                     border_gp = gpar(col = "NA"))

heat_female <- Heatmap(top20_gene_dat_female,cluster_rows = T,
                      cluster_columns = T,
                      top_annotation = ann_female,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                      col = colorRampPalette(my_color)(100),
                      na_col = "white",width =unit(8,"cm"),
                     height = unit(10,"cm"),
                     show_heatmap_legend = TRUE,
                     border_gp = gpar(col = "NA"))

p <- heat_male + heat_female
p
# pdf(paste0(fig_dir,"fig4f_top_20_gene_heatmap_20250328.pdf"),width = 16,height = 10)
# draw(p, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
# dev.off()

Heatmap(top20_gene_dat,name="top_20",
                      cluster_rows = T,
                      cluster_columns = T,
                     # right_annotation = row_annots,
                      top_annotation = ann_heatmap_df,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                      col = colorRampPalette(my_color)(100),
                      na_col = "white",width =unit(8,"cm"),
                     height = unit(10,"cm"),
                      # column_split = annotation_cols$Gender,
                     show_heatmap_legend = TRUE,
                     border_gp = gpar(col = "NA"))
```
## Figure S11

```{r,fig.width=16,fig.height=10}
heat_male <- Heatmap(top20_30_gene_dat_male,cluster_rows = T,
                      cluster_columns = T,
                      top_annotation = ann_male,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                      col = colorRampPalette(my_color)(100),
                      na_col = "white",width =unit(8,"cm"),
                     height = unit(10,"cm"),
                     show_heatmap_legend = FALSE,
                     border_gp = gpar(col = "NA"))

heat_female <- Heatmap(top20_30_gene_dat_female,cluster_rows = T,
                      cluster_columns = T,
                      top_annotation = ann_female,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                      col = colorRampPalette(my_color)(100),
                      na_col = "white",width =unit(8,"cm"),
                     height = unit(10,"cm"),
                     show_heatmap_legend = TRUE,
                     border_gp = gpar(col = "NA"))
p <- heat_male + heat_female
p
# pdf(paste0(fig_dir,"figS8_top_20_30_gene_heatmap_20250328.pdf"),width = 16,height = 10)
# draw(p, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")
# dev.off()
Heatmap(top20_30_gene_dat,
                      cluster_rows = T,
                      cluster_columns = T,
                     # right_annotation = row_annots,
                      top_annotation = ann_heatmap_df,
                      show_column_names = F,
                     show_row_dend = F,
                     show_row_names = T,
                     row_title = NULL,
                      col = colorRampPalette(my_color)(100),
                      na_col = "white",width =unit(10,"cm"),
                     height = unit(10,"cm"),
                      # column_split = annotation_cols$Gender,
                     show_heatmap_legend = TRUE,
                     border_gp = gpar(col = "NA"))
```
## Figure 3G  Plot a violin plot for all associated DMR regions between GC and GP genes.

```{r,echo=FALSE}
## Import data
sample_list <- meta_df$sample_list
dat_proj <- "~/methylation/project/EM_seq/analysis/DMR_analysis/data/filter_sample/dmr_file_delta0.15/"
## change data format
sample_DMR_meth <- read.table(paste0(dat_proj,"dmrs_region_list.bed"),col.names = "chr_region")
sample_DMR_cov <- read.table(paste0(dat_proj,"dmrs_region_list.bed"),col.names = "chr_region")

for (i_sample in sample_list){
  a <- read.table(paste0(dat_proj,i_sample,"_dmr_meth.bed"),col.name = c("chromosome","start","end","mod","unmod","meth_level"))
 sample_DMR_meth[[i_sample]] <- a$meth_level
 sample_DMR_cov[[i_sample]] <- a$mod+a$unmod
}
dmr_ann <- read.csv("~/methylation/project/EM_seq/analysis/DMR_analysis/code/gene_anlyais/meth_merge_change.csv")
## change data format
dmr_ann[, c(1:5)] <- lapply(dmr_ann[,c(1:5)], as.character)
dmr_ann$gene_region <-  apply(dmr_ann[, c(1:5)], 1, function(x) paste(x, collapse = "_"))
dmr_ann$chr_region <- apply(dmr_ann[, c(1:3)], 1, function(x) paste(x, collapse = "_"))
dmr_ann_df <- dmr_ann[,c(45,44,6:43)]
## merge smp_meth_data into gene annotation
merge_smp_meth_ann <- merge(dmr_ann_df,sample_DMR_meth,by="chr_region")
smp_meth_ann <- merge_smp_meth_ann[,c(2,41:78)]

## merge smp_cov_data into gene annotation
merge_smp_cov_ann <- merge(dmr_ann_df,sample_DMR_cov,by="chr_region")
smp_cov_ann <- merge_smp_cov_ann[,c(2,41:78)]

## sample_meth
data <- smp_meth_ann
rownames(data) <- as.character(data$gene_region)
data <- data[, -1]
tr_data <- t(data)
rownames(tr_data) <- colnames(data)
colnames(tr_data) <- rownames(data)
tr_data_meth <- as.data.frame(tr_data)

## sample_cov
data_cov <- smp_cov_ann
rownames(data_cov) <- as.character(data_cov$gene_region)
data_cov <- data_cov[, -1]
tr_data_co <- t(data_cov)
rownames(tr_data_co) <- colnames(data_cov)
colnames(tr_data_co) <- rownames(data_cov)
tr_data_cov<- as.data.frame(tr_data_co)

## Write a loop to do a plot 
common_cols <- intersect(names(tr_data_cov), names(tr_data_meth))
count=0
p_adjust <- list()
p_value <- list()
## 
for (col in common_cols) {
  df <- data.frame(x = tr_data_cov[[col]], y = tr_data_meth[[col]]) 
  df$status <- meta_df$sc_type%>%as.factor()
  df$smp <- str_sub(rownames(tr_data_cov),1,4)
  df <- na.omit(df)
    wilcox_test <- wilcox.test(df$y[df$status=="GP"],
                               df$y[df$status=="GC"],paired=F)
    
    p.value<- wilcox_test$p.value
    p_value[[col]] <- p.value
    
  if(!is.na(p_value)&&p_value< 0.05&&length(p_value)!=0){ 
        p1 <- ggplot(df, aes(x = status, y = y)) +
              geom_violin(aes(fill=status,alpha=1),scale = "width") +
              scale_fill_manual(values =c("#FC9595", "#C4E3FF"))+
              geom_boxplot(aes(fill=status),
                            outlier.shape = NA,lwd=0.4,width=0.05)+
              scale_fill_manual(values =c("#FB4F4F", "#93CDFF"),guide = 'none')+
              labs(y ="Methylation level", x = paste("y_", col), 
                   title = paste0("p = ", round(p_value, 10)))+
              theme_classic()+
              theme(axis.text.x = element_text(size = 15),
                    axis.text.y = element_text(size = 15),
                    axis.text = element_text(size = 15,colour = "black"),
                    axis.line.x = element_line(size = 1,linetype = 1,colour = "black"),
                    axis.line.y = element_line(size = 1,linetype = 1,colour = "black"),
                    axis.title = element_text(size = 15,colour = "black"),
                    legend.position="none")
        # ggsave(plot = p1,filename =paste0(fig_proj,"fig4C_gene_violin_plot_20240902/","meth_",
        #         col,"_violin_plot_20240902.pdf"),width = 6,height = 4,dpi = 300
        print(p1)
         }else{
             print(paste0(col,"_wilcox p_value less than 0.05,and the p is ",p_value))
            count=count+1 }
          print(count)
 }

```
## Fig 3I

```{r}
dmr_450k <- read_excel(paste0(dat_dir,"TCGA_Mutated_Genes.xlsx"),sheet = "TCGA_HM450_Mutated_Genes")
dmr_GC <- read_excel(paste0(dat_dir,"GI_dmr_anno.xlsx"),sheet = "GC")
homer_dmr <- read.csv(paste0(dat_dir,"fil_dmr_meth_gene.csv"))
D_45 <- dmr_450k$Gene
D_GC <- dmr_GC$Gene.Name%>% na.omit()
H_dmr <- homer_dmr$Genes
dmr_list <- list(D_45,H_dmr,D_GC)
names(dmr_list) <- c("D_45","H_dmr","D_GC")

# pdf(paste0(fig_dir,"filter_dmr_GC_vennplot.pdf"))
venn.plot1 <- venn.diagram(dmr_list[2:3],NULL,
             fill=c("#377EB8","#99D8C9"),
             alph=c(0.5,0.5),
             resolution = 300,
              cex=3,
             cat.fontface=3,
             category.names = c("DMR","GC"))
grid.draw(venn.plot1)
# dev.off()      
```
## technical

```{r}
sessionInfo()
```





